---
title: "Disease Outbreaks: Ben's Work"
author: "bml72"
date: "April 6, 2020"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, results = 'hide', message = F)

options(scientific=T, digits = 3) 
```

```{r basic, include=F}
loadPkg = function(pkg, character.only = FALSE) { 
  if (!character.only) { pkg <- as.character(substitute(pkg)) }
  pkg <- ifelse(!character.only, as.character(substitute(pkg)) , pkg)  
  if (!require(pkg,character.only=T, quietly =T)) {  install.packages(substitute(pkg),dep=T); if(!require(pkg,character.only=T)) stop("Package not found") } 
}
loadPkg(knitr)

unloadPkg = function(pkg, character.only = FALSE) { 
  if(!character.only) { pkg <- as.character(substitute(pkg)) } 
  search_item <- paste("package", pkg,sep = ":") 
  while(search_item %in% search()) { detach(search_item, unload = TRUE, character.only = TRUE) } 
}
```

## Pre-processing

Here, the data is read in as 'outbreaks.'
```{r reading in data, include = T}
outbreaks <- read.csv('outbreaks.csv')

str(outbreaks)

summary(outbreaks)
```

#### 'NA' Data

After skimming through the data set, I noticed that several predictors contained a large number of missing values. It is important to note that the missing 
values are denoted as '' instead of 'NA.' The largest occurances of missing values were in Serotype/Genotype (15,212 points, or 79.56% of datatset) and 
Ingredient (17,243 points, or 90.18% of dataset). Food contained 8,963 missing values (46.88% of dataset). Species and Status also contained a respectable 
number of missing values (6,619 points, or 34.62% of dataset).
```{r exploring the data, include = T}
year_na <- subset(outbreaks, Year == '') # 0 missing values

month_na <- subset(outbreaks, Month == '') # 0 missing values

location_na <- subset(outbreaks, Location == '') # 2,166 missing values (11.32% of dataset)

food_na <- subset(outbreaks, Food == '') # 8,963 missing values (46.88% of dataset)

ingredient_na <- subset(outbreaks, Ingredient == '') # 17,243 missing values (90.18% of dataset)

species_na <- subset(outbreaks, Species == '') # 6,619 missing values (34.62% of dataset)

serotype.genotype_na <- subset(outbreaks, Serotype.Genotype == '') # 15,212 missing values (79.56% of dataset)

status_na <- subset(outbreaks, Status == '') # 6,619 missing values (34.62% of dataset)

illnesses_na <- subset(outbreaks, Illnesses == '') # 0 missing values

hospitalizations_na <- subset(outbreaks, Hospitalizations == '') # 0 missing values

fatalities_na <- subset(outbreaks, Fatalities == '') # 0 missing values
```

#### Outlier Discovery

After constructing boxplots on the numerical variables (excluding year), outliers were discovered in illnesses, hospitalizations and fatalities.

• Illnesses: with an IQR of 24, there are technically 1,782 outliers. 

• Hospitalizations: with an IQR of 1.5, there are technically 1,325 outliers. 

• Fatalities: with an IQR of 0, there are technically 176 outliers. 

However, these outliers are significant and should not be excluded. Outliers can help explain what species of foodborne illnesses are the 
most destructive, or what predictors are especially harmful (i.e. State, Location). For example, in the case of illnesses the largest 
observation was recorded in a multistate area at a restaurant/private home residence/catering service and came from salmonella. Perhaps 
the location has something to do with why so many people got sick. Or, salmonella could be an especially dangerous disease. Excluding this 
outlier could produce harmful effects and cloud judgement on what factors predict the causative agent in foodborne disease outbreaks. 
```{r graphs, include = T}
illnesses_bp <- boxplot(outbreaks$Illnesses, main = "Boxplot of Illnesses", ylab = "Number of Illnesses")

outliers_illnesses <- subset(outbreaks, Illnesses > 43)

hospitalizations_bp <- boxplot(outbreaks$Hospitalizations, main = "Boxplot of Hospitalizations", ylab = "Number of Hospitalizations")

outliers_hospitalizations <- subset(outbreaks, Hospitalizations > 2.5)

fatalities_bp <- boxplot(outbreaks$Fatalities, main = "Boxplot of Fatalities", ylab = "Number of Fatalities")

outliers_fatalities <- subset(outbreaks, Fatalities > 0)
```

#### Excluding Insignificant Features

Because the majority of values for the `Ingredient`, `Serotype/Genotype` and `Status` features are missing, we are excluding them from our data set. 
```{r exclude NAs, include = T}
outbreaks <- outbreaks[-c(6, 8, 9)]

str(outbreaks)
```







```{r rf, include = T}
loadPkg('randomForest')

set.seed(100)

# Training conducted using 3:1 training to testing ratio (without replacement)
train <- sample(nrow(model_data), 0.75*nrow(model_data), replace = FALSE)

train_set <- model_data[train,]
valid_set <- model_data[-train,]

# summary(train_set)

rf <- randomForest(filterSpecies ~., data = train_set, mtry = 2, importance = TRUE)

rf

```





























