---
title: "DiseaseOutbreaks"
author: "Wasted Potential: Sneha, RyeAnne, Yuqi, Ben"
date: "4/2/2020"
output:
  html_document:
    code_folding: hide
    # number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = 'markup', message = F)
knitr::opts_chunk$set(warning = F, results = 'hide', message = F)
# knitr::opts_chunk$set(include = F)
# knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r basic, include=F}
# use this function to conveniently load libraries and work smoothly with knitting
# can add quietly=T option to the require() function
loadPkg = function(pkg, character.only = FALSE) { 
  if (!character.only) { pkg <- as.character(substitute(pkg)) }
  pkg <- ifelse(!character.only, as.character(substitute(pkg)) , pkg)  
  if (!require(pkg,character.only=T, quietly =T)) {  install.packages(substitute(pkg),dep=T); if(!require(pkg,character.only=T)) stop("Package not found") } 
}
loadPkg(knitr)
loadPkg(tidyverse)
loadPkg(ggplot2)
loadPkg(stringr)
loadPkg(maps)
loadPkg(mapdata)
loadPkg(usmap)


# unload/detact package when done using it
unloadPkg = function(pkg, character.only = FALSE) { 
  if(!character.only) { pkg <- as.character(substitute(pkg)) } 
  search_item <- paste("package", pkg,sep = ":") 
  while(search_item %in% search()) { detach(search_item, unload = TRUE, character.only = TRUE) } 
}
```


## Pre-processing

Here, the data is read in as 'outbreaks.'
```{r reading in data, include = T}
outbreaks <- read.csv('outbreaks.csv')

str(outbreaks)

summary(outbreaks)
```

# Summary of Data
```{r}
outbreak <- read.csv("outbreaks.csv")
summary(outbreak)
str(outbreak)
```

#### 'NA' Data

After skimming through the data set, I noticed that several predictors contained a large number of missing values. It is important to note that the missing 
values are denoted as '' instead of 'NA.' The largest occurances of missing values were in Serotype/Genotype (15,212 points, or 79.56% of datatset) and 
Ingredient (17,243 points, or 90.18% of dataset). Food contained 8,963 missing values (46.88% of dataset). Species and Status also contained a respectable 
number of missing values (6,619 points, or 34.62% of dataset).
```{r exploring the data, include = T}
year_na <- subset(outbreaks, Year == '') # 0 missing values

month_na <- subset(outbreaks, Month == '') # 0 missing values

location_na <- subset(outbreaks, Location == '') # 2,166 missing values (11.32% of dataset)

food_na <- subset(outbreaks, Food == '') # 8,963 missing values (46.88% of dataset)

ingredient_na <- subset(outbreaks, Ingredient == '') # 17,243 missing values (90.18% of dataset)

species_na <- subset(outbreaks, Species == '') # 6,619 missing values (34.62% of dataset)

serotype.genotype_na <- subset(outbreaks, Serotype.Genotype == '') # 15,212 missing values (79.56% of dataset)

status_na <- subset(outbreaks, Status == '') # 6,619 missing values (34.62% of dataset)

illnesses_na <- subset(outbreaks, Illnesses == '') # 0 missing values

hospitalizations_na <- subset(outbreaks, Hospitalizations == '') # 0 missing values

fatalities_na <- subset(outbreaks, Fatalities == '') # 0 missing values
```

#### Outlier Discovery

After constructing boxplots on the numerical variables (excluding year), outliers were discovered in illnesses, hospitalizations and fatalities.

• Illnesses: with an IQR of 24, there are technically 1,782 outliers. 

• Hospitalizations: with an IQR of 1.5, there are technically 1,325 outliers. 

• Fatalities: with an IQR of 0, there are technically 176 outliers. 

However, these outliers are significant and should not be excluded. Outliers can help explain what species of foodborne illnesses are the 
most destructive, or what predictors are especially harmful (i.e. State, Location). For example, in the case of illnesses the largest 
observation was recorded in a multistate area at a restaurant/private home residence/cateering service and came from salmonella. Perhaps 
the location has something to do with why so many people got sick. Or, salmonella could be an especially dangerous disease. Excluding this 
outlier could produce harmful effects and cloud judgement on what factors predict the causitive agent in foodborne disease outbreaks. 
```{r graphs, include = T}
illnesses_bp <- boxplot(outbreaks$Illnesses, main = "Boxplot of Illnesses", ylab = "Number of Illnesses")

outliers_illnesses <- subset(outbreaks, Illnesses > 43)

hospitalizations_bp <- boxplot(outbreaks$Hospitalizations, main = "Boxplot of Hospitalizations", ylab = "Number of Hospitalizations")

outliers_hospitalizations <- subset(outbreaks, Hospitalizations > 2.5)

fatalities_bp <- boxplot(outbreaks$Fatalities, main = "Boxplot of Fatalities", ylab = "Number of Fatalities")

outliers_fatalities <- subset(outbreaks, Fatalities > 0)
```

# EDA

### Illnesses across the states change in different years. 

```{r}
levels(outbreak$State)[levels(outbreak$State)=="Washington DC"] <- "District of Columbia"

# In 1998
ill1998 <- outbreak %>% filter(Year == 1998) %>%
  group_by(State) %>% 
  summarise(illness = mean(Illnesses, na.rm=T)) %>%
  left_join(x=statepop, by=c("full"="State")) %>%
  select(fips, abbr, full, illness)
plot_usmap(data=ill1998, values="illness", color=1, labels = T) +
  scale_fill_continuous(name="Illnesses",
                        low = "yellow", high = "red",
                        label=scales::comma) + 
  ggtitle(paste0("Illnesses by state in ", 1998)) +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, size=14))

# In 2004
ill2004 <- outbreak %>% filter(Year == 2004) %>%
  group_by(State) %>% 
  summarise(illness = mean(Illnesses, na.rm=T)) %>%
  left_join(x=statepop, by=c("full"="State")) %>%
  select(fips, abbr, full, illness)
plot_usmap(data=ill2004, values="illness", color=1, labels = T) +
  scale_fill_continuous(name="Illnesses",
                        low = "yellow", high = "red",
                        label=scales::comma) + 
  ggtitle(paste0("Illnesses by state in ", 2004)) +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, size=14))

# In 2010 
ill2010 <- outbreak %>% filter(Year == 2010) %>%
  group_by(State) %>% 
  summarise(illness = mean(Illnesses, na.rm=T)) %>%
  left_join(x=statepop, by=c("full"="State")) %>%
  select(fips, abbr, full, illness)
plot_usmap(data=ill2010, values="illness", color=1, labels = T) +
  scale_fill_continuous(name="Illnesses",
                        low = "yellow", high = "red",
                        label=scales::comma) + 
  ggtitle(paste0("Illnesses by state in ", 2010)) +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, size=14))

# In 2015
ill2015 <- outbreak %>% filter(Year == 2015) %>%
  group_by(State) %>% 
  summarise(illness = mean(Illnesses, na.rm=T)) %>%
  left_join(x=statepop, by=c("full"="State")) %>%
  select(fips, abbr, full, illness)
plot_usmap(data=ill2015, values="illness", color=1, labels = T) +
  scale_fill_continuous(name="Illnesses",
                        low = "yellow", high = "red",
                        label=scales::comma) + 
  ggtitle(paste0("Illnesses by state in ", 2015)) +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, size=14))

```


### The general trend of foodborne illnesses over the years and months

```{r}
# By year
outbreak %>% group_by(Year) %>%
  summarise(Illness = sum(Illnesses, na.rm=T)) %>%
  ggplot(aes(x = Year, y = Illness)) +
  geom_bar(stat = "identity", fill = "cadetblue3") +
  geom_text(aes(label = Illness), vjust = -0.4, color = "black",size=3) +
  ggtitle("Foodborne Illnesses change By Year") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "gray96"),
        axis.title.x=element_blank(),
        plot.title = element_text(hjust = 0.5, size=14))

# By Month
outbreak_month <- outbreak %>% 
  group_by(Month) %>%
  summarise(Illness = sum(Illnesses, na.rm=T)) 
outbreak_month$Month <- substr(outbreak_month$Month, 1, 3)
outbreak_month$Month <- factor(outbreak_month$Month, month.abb, ordered=TRUE)

outbreak_month %>% 
  ggplot(aes(x = Month, y = Illness)) +
  geom_bar(stat = "identity", fill = "cadetblue3") +
  geom_text(aes(label = Illness), vjust = -0.4, color = "black",size=3) +
  ggtitle("Foodborne Illnesses change By Month") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "gray96"),
        axis.title.x=element_blank(),
        plot.title = element_text(hjust = 0.5, size=14))

```


### Time series plot of illnesses caused by top4 species

```{r}
outbreak %>% filter(Species %in% c(
  "Norovirus genogroup I",
  "Salmonella enterica",
  "Norovirus genogroup II",
  "Clostridium perfringens"
)) %>%
  group_by(Year, Species) %>%
  summarise(Illnesses = sum(Illnesses, na.rm=T)) %>% 
  ggplot(aes(x=Year, y=Illnesses, lty=Species, col=Species)) + 
  geom_line()
```

# Model Building

## K-Nearest Neighbors

##Decision Trees

### Basic Decision Tree

### Random Forest

### Extreme Gradient Boost Tree